---
title: "Bioconductor package licenses"
format: 
    html:
        css: styles.css
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

# Get Bioc package list

```{r, include = FALSE}
options(width = 120)
```

```{r}
suppressPackageStartupMessages({
    library(BiocPkgTools)
    library(dplyr)
    library(DT)
})

BiocManager::version()
pkg <- biocPkgList(repo = "BioCsoft")
# pkg <- biocPkgList(repo = c("BioCsoft", "BioCann", "BioCexp", "BioCworkflows"))
```

# Get licenses

```{r}
sort(table(pkg$License, useNA = "ifany"))
```

# Clean up

```{r}
## Get unique licenses
lic <- unique(pkg$License)
alres <- lapply(structure(lic, names = lic), tools:::analyze_license)

## Get information about the licenses
alres <- do.call(dplyr::bind_rows, lapply(names(alres), function(nm) {
    a <- alres[[nm]]
    data.frame(License = nm,
               LicenseStandardizable = a$is_standardizable,
               LicenseVerified = a$is_verified,
               LicenseStandardization = a$standardization,
               LicenseCanonical = a$is_canonical,
               LicenseFOSS = a$is_FOSS,
               LicenseRestrictsUse = a$restricts_use)
}))

## Add additional column removing the "+/| file LICENSE" part (if we want to 
## assume that this is compatible with the main license indication)
alres$LicenseStandardizationShort <- gsub("[ ]*\\+*\\|*[ ]*file LICEN[SC]E",
                                          "", alres$LicenseStandardization)

dim(alres)
length(unique(alres$LicenseStandardizationShort))
DT::datatable(alres,
              extensions = "FixedColumns",
              rownames = FALSE,
              options = list(scrollX = TRUE, 
                             fixedColumns = list(leftColumns = 1),
                             search = list(regex = FALSE, caseInsensitive = TRUE),
                             pageLength = 20))
```


```{r}
## Add license columns to pkg list
pkg <- pkg %>% 
    dplyr::select(Package, Maintainer, License, hasLICENSE) %>%
    dplyr::left_join(alres, by = "License")
dim(pkg)
DT::datatable(pkg,
              extensions = "FixedColumns",
              rownames = FALSE,
              options = list(scrollX = TRUE, 
                             fixedColumns = list(leftColumns = 1),
                             search = list(regex = FALSE, caseInsensitive = TRUE),
                             pageLength = 20))
```

# Tabulate licenses

```{r}
sort(table(pkg$LicenseStandardizationShort, useNA = "ifany"))
```

# Some cases that may need attention

## Creative Commons licenses

Creative Commons recommends not using their licenses for software (but may be suitable for data packages): [https://creativecommons.org/faq/#can-i-apply-a-creative-commons-license-to-software](https://creativecommons.org/faq/#can-i-apply-a-creative-commons-license-to-software). Note though that the CC licenses are among the 'standard' open source licenses listed in the R license database: [https://svn.r-project.org/R/trunk/share/licenses/license.db](https://svn.r-project.org/R/trunk/share/licenses/license.db).

```{r}
DT::datatable(subset(pkg, grepl("CC|Creative Commons", LicenseStandardizationShort)),
              extensions = "FixedColumns",
              rownames = FALSE,
              options = list(scrollX = TRUE, 
                             fixedColumns = list(leftColumns = 1),
                             search = list(regex = FALSE, caseInsensitive = TRUE),
                             pageLength = 20))
```

## Packages that have just a 'file LICENSE'

These need to be checked manually.

```{r}
.makeLinkToLicense <- function(pkg, text) {
    sprintf('<a href="%s" target="_blank"> %s</a>',
            paste0("https://bioconductor.org/packages/devel/bioc/licenses/", 
                   pkg, "/LICENSE"), 
            text)
}
DT::datatable(subset(pkg, LicenseStandardizationShort == "") %>% 
                  dplyr::mutate(License = vapply(seq_along(License), function(i) {
                      .makeLinkToLicense(Package[i], License[i])
                  }, "NA")),
              extensions = "FixedColumns",
              rownames = FALSE,
              escape = FALSE,
              options = list(scrollX = TRUE, 
                             fixedColumns = list(leftColumns = 1),
                             search = list(regex = FALSE, caseInsensitive = TRUE),
                             pageLength = 20))
```

## Packages that have license "Unlimited"

```{r}
DT::datatable(subset(pkg, LicenseStandardizationShort == "Unlimited"),
              extensions = "FixedColumns",
              rownames = FALSE,
              options = list(scrollX = TRUE, 
                             fixedColumns = list(leftColumns = 1),
                             search = list(regex = FALSE, caseInsensitive = TRUE),
                             pageLength = 20))
```

## Packages that have licenses that can not be standardized

Some of these may be false positives. 

```{r}
DT::datatable(subset(pkg, is.na(LicenseStandardizationShort)),
              extensions = "FixedColumns",
              rownames = FALSE,
              options = list(scrollX = TRUE, 
                             fixedColumns = list(leftColumns = 1),
                             search = list(regex = FALSE, caseInsensitive = TRUE),
                             pageLength = 20))
```

The rest of the licenses appear to be approved open source licenses ([https://opensource.org/licenses/alphabetical](https://opensource.org/licenses/alphabetical)), although not always properly specified with a version etc. 

# Session info

```{r}
sessionInfo()
```

